Place of death in NRS COVID-19 deaths

### 2021-04-16 covid_nrs_place_of_death_visualisations.Rmd
Split the .Rmd into one to deal with data and another to deal with visualisations for easier working

# Set up SPARQL query

Note: not needed if using `opendatascot`
To access the latest NRS data using their API, we first setup a SPARQL query.

```{r}
# endpoint <- "http://statistics.gov.scot/sparql.json"
endpoint <- "http://statistics.gov.scot/sparql"
```

@@

# Context of lockdowns

A recent timeline of COVID-19 events in Scotland is [available here](https://spice-spotlight.scot/2021/01/29/timeline-of-coronavirus-covid-19-in-scotland/). Including these might help in the interpreration fo the figures, though the figures are already pretty busy. This is a potential future TODO!

# Load info about latest data

A separate script checks for updated data and downloads it, also updating these files with the locations of the downloaded data:

```{r}
file_data_sources <- "./table_of_data_sources.csv"
file_historical_data_sources <- "./table_of_historical_data_sources.csv"
file_case_trends_data_sources <- "./table_of_case_trends_data_sources.csv"

table_of_data_sources <- read_csv(file = file_data_sources)
table_of_historical_data_sources <- read_csv(file = file_historical_data_sources)
table_of_case_trends_data_sources <- read_csv(file = file_case_trends_data_sources)

table_of_data_sources %>%
  knitr::kable(caption = "Latest data used in this report.")
```

## Set up output directory

Because data are updated on a weekly basis, we'll have a separate directory for every week the data are updated:

```{r}
latest_data_modified_date <- strftime(max(table_of_data_sources$last_modified), format="%F")

dir_outputs <- paste0("./outputs/",latest_data_modified_date)
if (!dir.exists(dir_outputs)) dir.create(dir_outputs)
```

# Import & wrangle data

We're importing two kinds of data:

1. Weekly deaths involving COVID-19 from Scottish gov't open data service, broken down by week & place of death (home or institution)
2. Historical deaths 2015-2019, grouped by week & place of death

For both of these, we'll also look at crosstabulated figures for:

* Health board
* Local Authority
* Sex
* Age
* ~~Ethnicity~~ *Note: not sure this is available*

There are two ways of obtaining the data:

1. ~~Using their SPARQL-based API ("data cube"), which can be accessed using the `opendatascot` package in R; the data can be seen on [the Scottish Gov't open data website](https://statistics.gov.scot/data/deaths-involving-coronavirus-covid-19)~~
    - Update: it's difficult to keep track of data updates using the API - it's simpler to query the .xlsx/.csv files (next bullet point) for when they were last modified and download the weekly update once it's been released!
2. There are a number of datasets, including the demographic/geographic crosstabulations available on the "Related Statistics" section of the [deaths involving COVID-19 website]( https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/general-publications/weekly-and-monthly-data-on-births-and-deaths/deaths-involving-coronavirus-covid-19-in-scotland), under the *Ad-hoc queries section*. They are updated weekly, and also contain historical figures.

Note that the [*About* tab of the "data cube" interface](https://statistics.gov.scot/resource?uri=http%3A%2F%2Fstatistics.gov.scot%2Fdata%2Fdeaths-involving-coronavirus-covid-19) says the following about Health Board & Local Authority figures:

>Health Board and Local Authority figures include non-residents. Deaths are allocated to areas based on the usual residence of the deceased. If the deceased was not a Scottish resident, the death is allocated to the area where the death occurred.

This means we'll need to download those figures from the website rather than using the API!

Note that the weekly deaths due to COVID-19 in 2020 & 2021 also include a weekly average 2015-2019, but for the purposes of showing the range of deaths (weekly minima & maxima), we need the weekly figures for individual years & the appropriate cross-tabulations.

```{r}
# TODO: remove this section, params not used anymore
# # "unpack" the data source tables
# table_of_data_sources <- params$data_sources
# table_of_historical_data_sources <- params$historical_data_sources
# table_of_case_trends_data_sources <- params$case_trends_data_sources
```

## Parsing week numbers

In the `opendatascot`-imported data, `refPeriod` refers to the week number (called `week of occurrence` in the `.xlsx`/`.csv`-imported data has), as defined by [ISO 8601](https://en.wikipedia.org/wiki/ISO_week_date), and can be computed from a date with `lubridate::isoweek()`; the first week of 2020 is the week starting with Monday 2019-12-30, ~~and the corresponding start dates for each week can be computed by adding weeks using `lubridate::dweeks()` to that date.~~ This is non-trivial, because the number of iso weeks in a year isn't constant. The `ISOweek::ISOweek2date()` function deals with this issue in a crude manner, and allows us to compute the date (w/c on date) from the year & week number. It relies on being provided a valid week number (some years have 52 and some have 53), but it's probably safe to assume that the NRS data has valid dates.

Invalid week numbers will produce a date in the next year: e.g. week 53 in 2021 would actually start in 2022, which makes it invalid. We can test for that by checking that the year provided is the same year in the date produced.

We'll also compute a 'run-over' week number of 2021 in case we plot both years on a contibuous X-axis - in this case, week 1 of 2021 follows week 53 of 2020, so we start counting from 54 in 2021.

```{r}
date_first_week_2020 <- ymd("2019-12-30")
ISOweek::ISOweek2date(weekdate = "2019-W53-1")
ISOweek::ISOweek2date(weekdate = "2020-W01-1")
ISOweek::ISOweek2date(weekdate = "2020-W52-1")
ISOweek::ISOweek2date(weekdate = "2020-W53-1")
ISOweek::ISOweek2date(weekdate = "2021-W01-1")
ISOweek::ISOweek2date(weekdate = "2021-W53-1")

compute_start_date_from_week_number <- function(week_number, year_number) {
  ## this assumes a valid week number! There are 52 or 53 weeks in a year
  isoweek_string <- glue::glue("{year_number}-W{str_pad(week_number, width=2, pad='0')}-1")
  # computed_dates <- ISOweek::ISOweek2date(isoweek_string)
  computed_dates <- rep(NA_Date_,times = length(week_number))
  computed_dates[which(!is.na(week_number))] <- ISOweek::ISOweek2date(isoweek_string[which(!is.na(week_number))])  # return NA instead of failing with ISOweek!
  ## invalid weeks will result in a computed date in a year after the requested year...
  ## a valid week can result in a date that starts in the year before!
  if (any(year_number<year(ymd(computed_dates)), na.rm = TRUE)) stop("invalid week number provided!")
  return(computed_dates)
}

tibble(
    week_number = c(1,2,3,52,53)
  ) %>%
  mutate(
    date = compute_start_date_from_week_number(week_number = week_number, year_number = 2020),
    iso_week_from_date = isoweek(date)
  ) %>% knitr::kable(caption = "Examples of computing start date for each week, and computing isoweek from the dates!")
```

## Consistent Place of death coding

Place of death was named differently between datasets (location/place) and the different places were also named slightly differently, so we find all of them and recode them to the same:

```{r}
order_of_place_of_death_levels <- c("Hospital", "Care home", "Home & other non-institution", "Other", "All")

recode_place_of_death <- function(data_tbl) {
  data_tbl %>%
    mutate(place_of_death = case_when(
      str_detect(place_of_death, pattern = regex(pattern = "all", ignore_case = TRUE)) ~ "All",
      str_detect(place_of_death, pattern = regex(pattern = "care.*home", ignore_case = TRUE)) ~ "Care home",
      str_detect(place_of_death, pattern = regex(pattern = "(non.*institut)|home", ignore_case = TRUE)) ~ "Home & other non-institution",
      str_detect(place_of_death, pattern = regex(pattern = "hospital", ignore_case = TRUE)) ~ "Hospital",
      str_detect(place_of_death, pattern = regex(pattern = "other", ignore_case = TRUE)) ~ "Other",
      TRUE ~ NA_character_
      ) %>%
        factor(x = ., levels = order_of_place_of_death_levels)
    )
}
```

## Consistent Cause of death coding

The `opendatascot`-imported data distinguishes "all causes" deaths from "covid-related" deaths - the non-covid-related deaths are the difference between those. In contrast, the other weekly datasets code cause of death as either covid-related or non-covid-related, in which case we need to compute all cause deaths as the sum of the two.

We'll change the wording of these to a consistent format so it's easier to compute consistent columns. The below function will also compute non-covid deaths for data where all deaths and covid deaths are provided, and all deaths where covid- and non-covid-related deaths are provided.

This involves pivotting the number of deaths variable into wide format based on the cause of death. Because of this, we need to deal with implicit missing values:

The first COVID-19-related deaths were reported from week 12 onwards. In some of the datasets, a missing/omitted value represents 0 cases in that category, when a 0 should have been recorded. E.g. deaths in "Other institutions" are very low, and if a record is missing for that category, we should assume 0 deaths.

On 3 Februrary 2021 as I was updating the script, the weekly figures started distinguishing cause of death: COVID-19 can be a contributory factor, or an underlying cause of death. For the purposes of this study, COVID-19 deaths are in the minority, and we will disregard this distinction for now.

```{r}
## helper function to make consistently coded cause of death
recode_cause_of_death <- function(data_tbl) {
  data_tbl %>%
    mutate(
      cause_of_death = case_when(
        cause_of_death == "all-causes" ~ "deaths_all_causes",
        cause_of_death == "all-causes-average-of-corresponding-week-over-previous-5-years" ~ "deaths_nrs_past_average_all_causes",
        cause_of_death == "covid-19-related" ~ "deaths_covid_related",
        cause_of_death == "COVID-19 mentioned" ~ "deaths_covid_related",
        cause_of_death == "COVID-19 contributory factor" ~ "deaths_covid_related",
        cause_of_death == "COVID-19 underlying cause" ~ "deaths_covid_related",
        cause_of_death == "Non-COVID-19" ~ "deaths_non_covid",
        TRUE ~ NA_character_
      )
    )
}

# TODO:  still too complicated to safely use as a function...
## find which combinations of geographic area, cause of death, and place of death are missing and set the number of deaths to 0 for them
# complete_weekly_deaths_with_implicit_zeros <- function(data_tbl) {
#   data_tbl %>%
#   tidyr::complete(
#       nesting(year, week_number, week_number_run_over, date_w_c),  # these are 'identifying' variables that shouldn't be combinatorily exploded
#       ref_area, cause_of_death, place_of_death,  # these are the variables we want to have all combinations of, and missing combinations are set to 0 deaths
#       fill = list(number_of_deaths = 0)
#       )
# }

## calculate number of deaths after recoding cause of death -
## recoding results in two entries for covid-related deaths, so we need to sum over them
## this is achieved by grouping over all variables other than number of deaths
recalculate_number_of_deaths_after_recoding_cause_of_death <- function(data_tbl) {
  data_tbl %>%
  group_by_all %>% 
  ungroup(number_of_deaths) %>%  # this is done by grouping over everything except the number of deaths
  summarise(number_of_deaths = sum(number_of_deaths), .groups = "drop")  # recompute the sum of deaths
}

## covid death reporting started in week 12, 2020 - before then it doesn't make sense to distunguish between covid and non-covid deaths, and only the all cause death figure should be kept
## by using the 'overrun' weeks we ensure that 2021 or later isn't affected
## note: we're keeping all cause deaths!
set_covid_and_non_covid_deaths_to_zero_before_reporting_started <- function(data_tbl) {
  data_tbl %>%
    mutate(
      deaths_non_covid = if_else(
        condition = week_number_run_over < 12,
        true = NA_real_,
        false = deaths_non_covid
      ),
      deaths_covid_related = if_else(
        condition = week_number_run_over < 12,
        true = NA_real_,
        false = deaths_covid_related
      )
    )
}

# TODO: delete below function, it's too complicated and I've broken it down to the individual functions
# recode_cause_of_death_and_compute_columns <- function(data_tbl) {
#   potential_grouping_variables <- c("ref_area","sex","age","ref_period")
#   grouping_variables <- potential_grouping_variables[which(potential_grouping_variables %in% names(data_tbl))]
#   
#   interim_data_tbl <- 
#     data_tbl %>%
#     recode_cause_of_death() %>%
#     tidyr::complete(
#       nesting(year, week_number, week_number_run_over, date_w_c),
#       !!!rlang::syms(grouping_variables),
#       cause_of_death, place_of_death,
#       fill = list(number_of_deaths = 0)
#       ) %>%
#     pivot_wider(names_from = "cause_of_death", values_from = "number_of_deaths")
#   
#   ## compute non-covid deaths from all deaths minus covid deaths
#   if (!"deaths_non_covid" %in% names(interim_data_tbl)) {
#     interim_data_tbl <-
#       interim_data_tbl %>%
#       mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related)
#   }
#   ## compute all deaths from non-covid deaths + covid deaths
#   if (!"deaths_all_causes" %in% names(interim_data_tbl)) {
#     interim_data_tbl <-
#       interim_data_tbl %>%
#       mutate(deaths_all_causes = deaths_non_covid + deaths_covid_related)
#   }
#   
#   interim_data_tbl <-
#     interim_data_tbl %>%  # set as missing covid- & non-covid-related deaths before week 12
#     mutate(
#       deaths_non_covid = if_else(
#         condition = week_number_run_over < 12,
#         true = NA_real_,
#         false = deaths_non_covid
#       ),
#       deaths_covid_related = if_else(
#         condition = week_number_run_over < 12,
#         true = NA_real_,
#         false = deaths_covid_related
#       )
#     )
#   
#   return(interim_data_tbl)
# }
```

## Consistent use of missing value

```{r}
replace_na_with_0_from_week_12 <- function(data_tbl) {
  data_tbl %>%
    mutate(
      value = if_else(
        condition = week_number_roll_over >=12 & is.na(number_of_deaths),
        true = 0,
        false = number_of_deaths
      )
    )
}
```

## Merging health boards

Smaller health boards were merged... TODO!

> "Note: Orkney, Shetland, & Western health boards were merged into 'Island HBs'; Borders & Dumfries and Galloway health boards were merged into 'B,D&G HB'"

## Load weekly COVID-19 related deaths

~~Using `opendatascot` to load weekly deaths the level of geography can be specified, so we can have more or less granular data~~ Choosing a more granular level than Scotland-wide data means place of death is collapsed to the 'all' category only, see note above! Even for the Scotland-wide data, place of death is collapsed to 'All' for the granular sex & age categories, actually.

I've also attempted loading the case counts using `ckanr` using the CKAN API from the opendata.nhs.scot website, but gave that up in favour of using their `.csv` links instead to download daily-updated data.

Note about weekly LA/HB data:

> From 14/12/2020 confirmed positive cases are now allocated to an area based on the postcode recorded in the testing system. Previously cases were allocated to local areas mainly based on their usual residential address recorded in the Community Health Index (CHI) database. This change will impact on newly reported cases on this day.

> Please note that the updated methodology for determining COVID-19 related ICU admissions has been applied to the historic data. The updated definition is as follows: A patient who has tested positive for COVID at any time in the 21 days prior to admission to ICU, or who have tested positive from the date of admission up to and including the date of ICU discharge.

> In addition, on 30th October it came to our attention that there is an ongoing issue when linking ICU data to the lab data for COVID-19 test results. Any COVID-19 positive patients with a missing CHI that had a first positive test in the community are unable to be linked to the ICU data. As a result the COVID-19 positive numbers could be underreported by a small number.

`crude_rate_positive`: Crude rate of total positive cases (cumulative) per 100,000 population.

### Define locations for weekly data fiels

```{r}
download_dir <- "./downloaded_data"
if (!dir.exists(download_dir)) dir.create(download_dir)  # create download dir if it doesn't exist

files_weekly_deaths <- set_names(x = as.list(table_of_data_sources$path_latest), nm = table_of_data_sources$short_name)
```

### Load  'main report' data

We are only loading deaths aggregated by cause from the 'main report' data! These are compiled in two sheets:

* Table 3 (2020)
* Table 3  (2021)

However, the number of spaces might differ! So we will extract the table names automatically.

The ICD-10 codes used for causes of death:

>4) The ICD 10 codes for disease categories are as follows:	
Cancer: C00-C97	
Dementia and Alzheimer's: F01, F03 and G30	
Circulatory: I00:I99	
Respiratory: J00-J99	
COVID-19: U07

The format is a bit complicated to read automatically; we will extract the number of weeks available in the data from the filename.

~~These data come chopped up into 4 sheets, 2 by 2 for each of year 2020/2021 and all deaths/covid-related deaths. They are named:

* Table 1 (2020)
* Table 1 (2021)
* Table 2 (2020)
* Table 2 (2021)

~~

```{r}
weeks_available_in_nrs_weekly_data_by_cause_of_death_2021 <-  # extract the maximum weeks available from the weekly-updated 2021 data
  files_weekly_deaths$main_report %>%
  str_extract(string = ., pattern = "week-\\d{2}") %>%  # extract eg "week-12"
  str_extract(string = ., pattern = "\\d{2}") %>%  # extract "12"
  as.integer()

## extract the sheet names for cause of death data
nrs_weekly_data_sheet_2020 <- 
  excel_sheets(files_weekly_deaths$main_report)[which(str_detect(tolower(excel_sheets(files_weekly_deaths$main_report)), pattern = "table 3.*2020"))]
nrs_weekly_data_sheet_2021 <- 
  excel_sheets(files_weekly_deaths$main_report)[which(str_detect(tolower(excel_sheets(files_weekly_deaths$main_report)), pattern = "table 3.*2021"))]

## construct column names for the data
headers_in_nrs_data_2020 <-
  c(
    "place_of_death",
    "cause_of_death",
    paste0("week_", 1:53),
    "blank_remove",  # this and the rest are irrelevant entries that we'll remove
    "remove_1",
    "remove_2"
  )
headers_in_nrs_data_2021 <-
  c(
    "place_of_death",
    "cause_of_death",
    paste0(
      "week_",
      1:weeks_available_in_nrs_weekly_data_by_cause_of_death_2021
    ),
    "blank_remove",  # this and the rest are irrelevant entries that we'll remove
    "remove_1"
  )

## extract data... this is dense but I've tried to annotate it clearly!
weekly_deaths_cause_of_death_before_split_2020 <-
  readxl::read_excel(files_weekly_deaths$main_report, sheet = nrs_weekly_data_sheet_2020, col_names = headers_in_nrs_data_2020, range = cell_rows(6:125)) %>%  # the data are inrows 6:125
  select(-matches("remove")) %>%  # remove irrelevant summary columns at the end
  mutate(across(.cols = matches("week"), .fns = as.numeric)) %>%  # remove erroneous "count1, count2, ..." entries due to superscript in previous column
  mutate(dataset = if_else(condition = str_detect(string = cause_of_death, pattern = regex(pattern = "^registered|^difference", ignore_case = TRUE)), true = cause_of_death, false = NA_character_)) %>%  # extract the Registered deaths: historical, 2020/2021, and Difference sections
  tidyr::fill(place_of_death, .direction = "down") %>%  # complete the entries for place of death from the "last observed"
  tidyr::fill(dataset, .direction = "down") %>%  # same as above but for "dataset" column
  drop_na(week_1) %>%  # remove blank "header" rows with no data recorded
  pivot_longer(cols = matches("week"), names_to = "week_number", values_to = "number_of_deaths", values_drop_na = TRUE) %>%  # convert to long data format
  mutate(
    week_number = as.integer(parse_number(week_number)),  # extract week number
    place_of_death =  # find best match in the reference names used in other datasets
      order_of_place_of_death_levels[stringdist::amatch(x = str_sub(place_of_death,1,4), table = str_sub(order_of_place_of_death_levels,1,4), method = "lv", maxDist = 5)]  # the reference is in the variable order_of_place_of_death_levels
    )

weekly_deaths_cause_of_death_before_split_2021 <-
  readxl::read_excel(files_weekly_deaths$main_report, sheet = nrs_weekly_data_sheet_2021, col_names = headers_in_nrs_data_2021, range = cell_rows(6:125)) %>%  # the data are inrows 6:125
  select(-matches("remove")) %>%  # remove irrelevant summary columns at the end
  mutate(across(.cols = matches("week"), .fns = as.numeric)) %>%  # remove erroneous "count1, count2, ..." entries due to superscript in previous column
  mutate(dataset = if_else(condition = str_detect(string = cause_of_death, pattern = regex(pattern = "^registered|^difference", ignore_case = TRUE)), true = cause_of_death, false = NA_character_)) %>%  # extract the Registered deaths: historical, 2020/2021, and Difference sections
  tidyr::fill(place_of_death, .direction = "down") %>%  # complete the entries for place of death from the "last observed"
  tidyr::fill(dataset, .direction = "down") %>%  # same as above but for "dataset" column
  drop_na(week_1) %>%  # remove blank "header" rows with no data recorded
  pivot_longer(cols = matches("week"), names_to = "week_number", values_to = "number_of_deaths", values_drop_na = TRUE) %>%  # convert to long data format
  mutate(
    week_number = as.integer(parse_number(week_number)),  # extract week number
    place_of_death =  # find best match in the reference names used in other datasets
      order_of_place_of_death_levels[stringdist::amatch(x = str_sub(place_of_death,1,4), table = str_sub(order_of_place_of_death_levels,1,4), method = "lv", maxDist = 5)]  # the reference is in the variable order_of_place_of_death_levels
    )

## compile the weekly data
weekly_deaths_cause_of_death <-
  bind_rows(
    weekly_deaths_cause_of_death_before_split_2020 %>% filter(!str_detect(tolower(dataset), pattern = "difference|average")),
    weekly_deaths_cause_of_death_before_split_2021 %>% filter(!str_detect(tolower(dataset), pattern = "difference|average"))
  ) %>%
  mutate(
    year = as.integer(parse_number(dataset)),
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
    ) %>%
  select(-dataset)

## compile historical data - note that it's already summarised as average over the 5 year period, so we don't have the raw data, nor do we have minima/maxima
historical_average_cause_of_death <-
  weekly_deaths_cause_of_death_before_split_2020 %>%  # only data from the 2020 sheet are used - the 2021 data is the same entries for historical deaths!
  filter(str_detect(tolower(dataset), pattern = "average")) %>%
  select(-dataset) %>%
  rename(deaths_mean = number_of_deaths) %>%
  mutate(deaths_min = NA_real_, deaths_max = NA_real_)
```

### Load 'ad-hoc' query data

These are data aggregated by sex, age, and region (HB/LA)!

```{r}
# checking out structure

# ods_all_datasets() %>% View()
# ods_structure("deaths")  # can we import all deaths-related data this way?
# ods_structure("coronavirus-covid-19-management-information")

# ods_dataset("coronavirus-covid-19-management-information", geography = "hb") %>%
#   filter(variable == "testing-new-cases-reported")
# 
# ods_dataset("coronavirus-covid-19-management-information", geography = "hb", variable = "testing-new-cases-reported")
# ods_dataset("coronavirus-covid-19-management-information", geography = "la", variable = "testing-new-cases-reported")
# ods_dataset("coronavirus-covid-19-management-information", variable = "testing-new-cases-reported")
# 
# ods_print_query(dataset = "coronavirus-covid-19-management-information", geography = "hb", variable = "testing-new-cases-reported") %>% dput
# ods_print_query(dataset = "coronavirus-covid-19-management-information", variable = "testing-new-cases-reported") %>% dput
#   
# modified_query <- "PREFIX qb: <http://purl.org/linked-data/cube#>                 PREFIX rdfs: <http://www.w3.org/2000/01/rdf-schema#>                 PREFIX xsd: <http://www.w3.org/2001/XMLSchema#> select   ?refArea ?refPeriod ?measureType ?variable ?value where { ?data qb:dataSet <http://statistics.gov.scot/data/coronavirus-covid-19-management-information>. ?data <http://purl.org/linked-data/sdmx/2009/dimension#refArea> ?refArea. ?data <http://purl.org/linked-data/sdmx/2009/dimension#refPeriod> ?refPeriod. ?data <http://purl.org/linked-data/cube#measureType> ?measureType. ?data <http://statistics.gov.scot/def/dimension/variable> ?variable. ?data ?measureType ?value.  filter (regex(str(?variable ), 'testing-new-cases-reported'))  } order by ?refPeriod ?refArea"
# 
# modified_query %>%
#   opendatascot:::ods_query_database(endpoint, query = .) %>%
#   opendatascot:::pre_process_data()
# 
# query <- ""

# 
# library(ckanr)
# ckanr_setup(url = "https://www.opendata.nhs.scot/")
# # package_list(as = "table")
# # tag_list(as = "table")
# package_search(q = 'daily', rows = 2, as = 'table') %>% str
# 
# 
# download_data_from_opendata_via_ckan_api <- function(resource_id) {
#   ckan <- ckanr::src_ckan("https://www.opendata.nhs.scot/")
#   dplyr::tbl(src = ckan$con, from = resource_id_daily_case_trends_la, ) %>% 
#   as_tibble()
# }
# 
# resource_id_daily_case_trends_la <- "427f9a25-db22-4014-a3bc-893b68243055"
# 
# resource_show(id = resource_id_daily_case_trends_la, as = "table")

# ods_print_query("deaths")

# ods_structure("deaths-involving-coronavirus-covid-19")
# ods_schemes("deaths-involving-coronavirus-covid-19")
# ods_print_query("deaths-involving-coronavirus-covid-19")
# ods_print_query("deaths-involving-coronavirus-covid-19", geography = "la")
# ods_find_lower_geographies("S92000003")  # Scotland's refArea = S92000003
# phsmethods::match_area("S92000003")  # "Scotland"

## helper function that links the 9-character geography code to its name
make_human_readable_ref_area <- function(data_tbl, ref_area = refArea) {
  enquoted_ref_area <- enquo(ref_area)
  data_tbl %>%
    mutate(
      !!enquoted_ref_area := phsmethods::match_area(!!enquoted_ref_area)
    )
}

# raw_covid_deaths <- list(
#   scotland = ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc"),
#   health_board = ods_dataset("deaths-involving-coronavirus-covid-19", geography = "hb"),
#   local_authority = ods_dataset("deaths-involving-coronavirus-covid-19",geography = "la")
# ) %>%
#   map(.x = ., .f = ~make_human_readable_ref_area(data_tbl = .x))

weekly_deaths_overall <-
  ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc") %>%
  make_human_readable_ref_area %>%
  clean_names() %>%
  ## below series of rename operations are a stupid way of dealing with the possibility of different variable name conventions in the ods_dataset data: sometimes they are in snake_case, and sometimes in camelCase, seemingly unpredictably
  rename_at(.vars = vars(matches("location", ignore.case = TRUE)), ~"place_of_death") %>%
  rename_at(.vars = vars(matches("period", ignore.case = TRUE)), ~"ref_period") %>%
  rename_at(.vars = vars(matches("area", ignore.case = TRUE)), ~"ref_area") %>%
  rename_at(.vars = vars(matches("measure", ignore.case = TRUE)), ~"measure_type") %>%
  rename_at(.vars = vars(matches("cause", ignore.case = TRUE)), ~"cause_of_death") %>%
  rename_at(.vars = vars(matches("sex", ignore.case = TRUE)), ~"sex") %>%
  rename_at(.vars = vars(matches("age", ignore.case = TRUE)), ~"age") %>%
  rename_at(.vars = vars(matches("value", ignore.case = TRUE)), ~"number_of_deaths") %>%  # rename so it's consistent with other data
  filter(!ref_period %in% as.character(2020:2021)) %>% # remove totals for 2020 & 2021, we only want the weekly figures
  mutate(
    year = parse_integer(str_extract(ref_period, pattern="^(2020|2021)")),
    week_number = parse_integer(str_remove(ref_period, pattern = "^(2020|2021)\\-")),
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    number_of_deaths = as.numeric(number_of_deaths),  # convert to numeric!
    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
  ) %>%
  select(-c(measure_type)) %>%  #  measure_type is count for all data
  recode_place_of_death() %>%
  recode_cause_of_death() %>%
  recalculate_number_of_deaths_after_recoding_cause_of_death %>%  # not needed here probably, but no harm in including it
  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%  # this dataset is well-formatted and has no implicit missing values!
  mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related)  # compute non-covid deaths from the other two figures - will be NA where either is missing

# weekly_deaths_sex_age_file <-
#   curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-age-sex.xlsx",
#                 destfile = tempfile(),
#                 quiet = FALSE)

weekly_deaths_sex_age <-  # structure here is a bit more complicated - headers in row 3, but sex & age headers in row 4
  readxl::read_excel(files_weekly_deaths$sex_age, sheet = 2, skip = 4, col_names = c("week_number","place_of_death","sex","age","cause_of_death","number_of_deaths"), guess_max = 1e5) %>%
  drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
  mutate(
    year = 2000L + parse_integer(str_sub(week_number, 1, 2)),
    week_number = parse_integer(str_sub(week_number, 4, 5)),
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
  ) %>%
  recode_cause_of_death() %>%
  recalculate_number_of_deaths_after_recoding_cause_of_death %>%  # recompute the sum of deaths
  tidyr::complete(  # complete implicit missing entries and make them 0
    nesting(week_number,week_number_run_over,date_w_c,year),
    age,sex,place_of_death,cause_of_death,
    fill = list(number_of_deaths = 0)
  ) %>%
  recode_place_of_death() %>%  # this needs to be done after recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%   # make separate columns for causes of death figures
  mutate(deaths_all_causes = deaths_non_covid + deaths_covid_related) %>%  # compute all cause deaths from the other two figures reported
  set_covid_and_non_covid_deaths_to_zero_before_reporting_started()

# 
# weekly_deaths_la_file <-
#   curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-council-area-location.xlsx",
#                 destfile = tempfile(),
#                 quiet = FALSE)

weekly_deaths_la <- 
  read_excel(files_weekly_deaths$la, sheet = 2, skip = 2) %>% 
  clean_names %>%
  rename(week_number = week_of_occurrence, ref_area = council, place_of_death = location_of_death, number_of_deaths = deaths) %>%
  filter(!is.na(number_of_deaths)) %>%  # remove blank rows & copyright entry
  mutate(
    year = 2000L + parse_integer(str_sub(week_number, 1, 2)),
    week_number = parse_integer(str_sub(week_number, 4, 5)),
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
  ) %>%
  recode_cause_of_death() %>%
  recalculate_number_of_deaths_after_recoding_cause_of_death %>%  # recompute the sum of deaths
  tidyr::complete(
      nesting(year, week_number, week_number_run_over, date_w_c),  # these are 'identifying' variables that shouldn't be combinatorily exploded
      ref_area, cause_of_death, place_of_death,  # these are the variables we want to have all combinations of, and missing combinations are set to 0 deaths
      fill = list(number_of_deaths = 0)
      ) %>%
  # complete_weekly_deaths_with_implicit_zeros() %>%
  recode_place_of_death() %>%  # this needs to be done before recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%   # make separate columns for causes of death figures
  mutate(deaths_all_causes = deaths_non_covid + deaths_covid_related) %>%  # compute all cause deaths from the other two figures reported
  set_covid_and_non_covid_deaths_to_zero_before_reporting_started()

# weekly_deaths_hb_file <-
#   curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-health-board-location.xlsx",
#                 destfile = tempfile(),
#                 quiet = FALSE)

weekly_deaths_hb <- 
  read_excel(files_weekly_deaths$hb, sheet = 2, skip = 2) %>% 
  clean_names %>%
  rename(week_number = week_of_occurrence, ref_area = health_board, place_of_death = location_of_death, number_of_deaths = deaths) %>%
  filter(!is.na(number_of_deaths)) %>%  # remove blank rows & copyright entry
  mutate(
    year = 2000L + parse_integer(str_sub(week_number, 1, 2)),
    week_number = parse_integer(str_sub(week_number, 4, 5)),
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
  ) %>%
  recode_cause_of_death() %>%
  recalculate_number_of_deaths_after_recoding_cause_of_death %>%  # recompute the sum of deaths
  tidyr::complete(
      nesting(year, week_number, week_number_run_over, date_w_c),  # these are 'identifying' variables that shouldn't be combinatorily exploded
      ref_area, cause_of_death, place_of_death,  # these are the variables we want to have all combinations of, and missing combinations are set to 0 deaths
      fill = list(number_of_deaths = 0)
      ) %>%
  # complete_weekly_deaths_with_implicit_zeros() %>%
  recode_place_of_death() %>%  # this needs to be done before recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%  # make separate columns for causes of death figures
  mutate(deaths_all_causes = deaths_non_covid + deaths_covid_related) %>%  # compute all cause deaths from the other two figures reported
  set_covid_and_non_covid_deaths_to_zero_before_reporting_started()

## compute weekly data for sex and age separately
## sum over sex or age, respectively
weekly_deaths_sex <-
  weekly_deaths_sex_age %>%
  filter(age!="all") %>%
  group_by(week_number, week_number_run_over, year, date_w_c, sex, place_of_death) %>%
  summarise(across(starts_with("deaths_"), ~sum(.x)), .groups = "drop")

weekly_deaths_age <-
  weekly_deaths_sex_age %>%
  filter(sex!="all") %>%
  group_by(week_number, week_number_run_over, year, date_w_c, age, place_of_death) %>%
  summarise(across(starts_with("deaths_"), ~sum(.x)), .groups = "drop")
```

### Checking week number coding

```{r}
weekly_deaths_sex_age %>% 
  count(year, date_w_c, week_number, week_number_run_over) %>% 
  slice(1:3, (nrow(.) - 3):nrow(.)) %>% 
  select(-n) %>% 
  knitr::kable(caption = "Checking the computation of dates from week numbers - note how the 1st week of 2020 started in 2019.")
```

## Load historical deaths data

The format of the .xlsx files is that sheet 1 is a table of contents, sheet 2 contains the data, and sheet 3 is a pivot table for subsetting data.

```{r, warning=FALSE}
files_historical_deaths <- set_names(x = as.list(table_of_historical_data_sources$path_latest), nm = table_of_historical_data_sources$short_name)

>>>>

@@

# Load most recent data

```{r}
load(file = "./workspace.RData")
```
